Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = print("world");
let x = print("hello");
...
\end{BVerbatim}
\end{equation}
In this case, substituting a term for a variable changes the order the I/O
actions are performed, which changes the program's observable behaviour. On the
other hand, suppose we have an effect which increments a counter. Then the same
transformation is sound, since \(x\) is used exactly once:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = rand();
let z = rand();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
can change this from a program which always returns true to one which can return
either true or false. On the other hand, if the effect is the possibility of
nontermination, then the substitution is sound. 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure expression as one which is all three of
central, copyable, and discardable.

@@

\TODO{terms vs expression/blocks/control-flow-graphs}

\TODO{standardize "branching to" wordmark? Is wordmark the word here?}

@@

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
  \(p = 0\) otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\).
  So the centrality \(p\) indicates whether it is safe to substitute this
  expression for a variable, and \(q\), plus the linearity of \(A\), control how
  many times that variable can occur: if \(\taff \subseteq q\) then the variable
  is called \textit{affine} or \textit{discardable} and it is permissible to
  leave it unused, whereas if \(\trel \subseteq q\) it is called
  \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
  times. An expression which is central, relevant, and affine is called
  \textit{pure}. The rules for this judgement are given in Figure
  \ref{fig:ssa-term-typing}.

@@

We can now get to the typing judgement for expressions in Figure
\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
for variables, with the requirement that, other than \(x\), all variables in
\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{expression}
\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
reflects that we're allowed to \textit{forget} capabilities, but not
\textit{add} them. The \rle{app} rule is also mostly standard: given an
instruction \(f\) which takes in an argument \(A\) and has a return type of
\(B\) and a \textit{central} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we
conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
\(f\) itself have centrality \(p\) and quantity \(q\), i.e. that \(f \in
\mc{I}_p^q(A, B)\). The \rle{pair} rule says that we can type a pair \((a, b)\)
in \(\Gamma\) if \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\)
and \(\Xi\) types \(b\); the quantity of a pair is simply the intersection of
the quantity of its components (this is expressed in the rule by allowing any
quantity which types \textit{both} components), and both components must be
central. Finally, we have the rules \rle{let} and \rle{let2}, corresponding to
let-bindings and destructuring let-bindings respectively. In both cases, we
split the context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is
used to type a central subterm \(a\), which is bound to variables \(x\) or
destructured to variables \(x, y\) respectively, in a potentially non-central
term \(e\) which is otherwise typed in \(\Delta\).

@@

The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
\rle{br} says that, given a pure term \(a: A\) (that is, central, relevant, and
affine), we can type a block \(\lbrb{\ell}{a}\) which branches to \(\ell\) with
argument \(a\), where \(\ms{L}\) contains the context \(\ell\) with an argument
type of \(A\) and an appropriate set of unused variables. The rule \rle{ite}
typechecks if-then-else statements by stating that, if \(\Gamma\) splits into
\(\Delta\) and \(\Xi\), where \(\Xi\) typechecks a pure, intuitionistic
discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both \(s, t\) to branch
into \(\ms{L}\), then the whole if-statement \(\ite{e}{s}{t}\) also typechecks
as branching into \(\ms{L}\) in constext \(\Gamma\). \rle{let-blk} and
\rle{let2-blk} are quite similar to their expression analogues, except that the
bound expression \(e\) is allowed to be impure (otherwise, programs could only
make use of pure functions!), and the remainder of the context plus the bound
variable is used to type a block rather than a term. Finally, probably the most
interesting rule is \rle{where}, which says that, given a block \(t\) which
branches to \(\ms{L}\) in context \(\Gamma\), and labels \(L\) mapping
\(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\) branches to \(\ms{K}\). Here, \(L\)
mapping \(\ms{L}\) to \(\ms{K}\) is defined using the typing rules \rle{nil-br},
which says that an empty control-flow graph simply forwards the input labels
directly to the output labels, and \rle{cons-br}, which says that if \(L\) maps
\(\ms{L}\) to either \(\ms{K}\) or the label \(\lbl{\ell}[\Gamma](A)\), and
\(t\) branches to \(\ms{L}\) in the context \(\Gamma, \thyp{x}{A}{}\), then
\(L\) maps \(\ms{L}\) to \(\ms{K}\), having (potentially recursively) defined
\(\lbl{\ell}\).

@@

Given such a substitution \(\gamma\), there are two ways to use it. The first is
standard capture-avoiding substitution, which we denote \([\gamma]e\) for
expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for CFGs, as is
standard. The second is to use the substitution to generate a set of
let-bindings for a context \(\Gamma\), written
\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
CFGs. In Section \ref{sec:semantics}, we will show that this is semantically
equivalent to regular capture-avoiding substitution, due to the fact that all
terms in a substitution are central \textit{and} that only affine computations
can be used in an expression substituting for an affine variable, and only
relevant computationbs can be used in an expression substituting for a relevant
variable.

@@

We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
relevant term becomes a relevant morphism, and so, in particular, a pure term
becomes a pure morphsim. Our rules for doing so are given in Figure
\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
\(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
"judgement"; this is justified by the coherence lemma:
\begin{lemma}[Coherence] \
  \begin{itemize}
    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
  \end{itemize}
\end{lemma}
Our treatment of \rle{var} is quite standard: the expression \(x\) types in
\(\Gamma\) if all other variables in \(\Gamma\) and can be discarded, in which
case the semantics is simply doing so, i.e.
\(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\), lifting to \(\mc{C}_0\) as necessary.
Similarly, to type a function application \rle{var}, we first take
\(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is guarnateed to be central, lift it
to the appropriate category \(\mc{C}_p\), and then simply postcompose it with
the denotation of \(f\). \rle{pair} is handled by composing the context split
\(\csplits{\Gamma}{\Delta}{\Xi}\) taking the tensor product of the components
\(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\), \(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is
well-defined since both components are central and \(\mc{C}_1\) is monoidal. We
then lift the result to the appropriate \(\mc{C}_p\). The semantics of
\rle{unit} is simply dropping the entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\),
while \rle{true}, and \rle{false} are each handled by first dropping the context
and then postcomposing with the appropriate constant; in all cases, this is
followed by lifting to \(\mc{C}_p\). Finally, \rle{let} and \rle{let2} are
handled by first splitting the context, then passing the right hand component to
the semantics of the bound expression \(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or
\(\dnt{\hasty{\Xi}{1}{a}{A \otimes B}{q}}\), and then, lifting to \(\mc{C}_p\)
as necessary, after reassociating with \(\alpha\), passing everything to the
inner expression \(\dnt{\hasty{\Delta, \thyp{x}{A}{q}}{p}{e}{B}{q}}\) or
\(\dnt{\hasty{\Delta, \thyp{x}{A}{q}, \thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that
this inner expression is not necessarily central, unlike the bound expression!

@@

Similarly to terms, \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) are
interpreted as \textit{impure} morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\):
here \(\Gamma\) represents all the variables live on entry to the block, while
\(\ms{L}\) represents all the external blocks which \(t\) might branch to, each
annotated with their live variables and their arguments; our rules are given in
Figure \ref{fig:ssa-block-semantics}. Our semantics for break statements
\(\lbrb{\ell}{a}\) is as follows: we begin by splitting the context into live
variables which pass into dominated blocks and variables used in the argument
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\). We then pass the latter into the
semantics of the argument (which is pure, and therefore must be lifted to
\(\mc{C}_0\)) \(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result and live
variables re-associated and passed into
\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target. The
semantics of if-statements are similarly given by first splitting the context
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) into variables live in the branches,
\(\Delta\), and variables appearing in the discriminant, \(\Xi\); the latter of
which are passed into the lifted pure morphism
\(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\). The "if" statement itself is then
simply an application of the distributor \(\delta: A \otimes (\mb{1} + \mb{1})
\to A + A\), with each branch passed to the semantics of its corresponding block
\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\), \(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The
semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
\rle{let} and \rle{let2} in the expression case, except that now the bound
expression \(a\) is allowed to be impure (and therefore in particular does not
need to be upgraded).

@@

Because we have a denotational semantics, it is easy to validate
equations between terms.

@@

To properly be able to reason about congruence and congruence-based
optimization, we need to introduce the concept of \textit{terms with holes},
which are defined using the grammar in Figure
\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
syntactic categories with \textit{holes} of the form \(\lhole{X}\),
\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
introduce the concept of a \textit{hole-context} \(H\), which types a hole as being
one of:
\begin{itemize}
  \item \todo{expression}
  \item \todo{block}
  \item \todo{control flow graph}
\end{itemize}
\TODO{hole typing rules for expressions, blocks, control-flow graphs; everything
else is the same}.

>>>>

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = print("world");
let x = print("hello");
...
\end{BVerbatim}
\end{equation}
In this case, substituting a term for a variable changes the order the I/O
actions are performed, which changes the program's observable behaviour. On the
other hand, suppose we have an effect which increments a counter. Then the same
transformation is sound, since \(x\) is used exactly once:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = rand();
let z = rand();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
can change this from a program which always returns true to one which can return
either true or false. On the other hand, if the effect is the possibility of
nontermination, then the substitution is sound. 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure term as one which is all three of
central, copyable, and discardable.

@@

@@

Our grammar proper splits program fragments into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the term \(a\)
  is of type \(A\) with \textit{centrality} \(p = 1\) if \textit{central}, or
  \(p = 0\) otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\).
  So the centrality \(p\) indicates whether it is safe to substitute this
  term for a variable, and \(q\), plus the linearity of \(A\), control how
  many times that variable can occur: if \(\taff \subseteq q\) then the variable
  is called \textit{affine} or \textit{discardable} and it is permissible to
  leave it unused, whereas if \(\trel \subseteq q\) it is called
  \textit{relevant} or \textit{duplicable} and it is permissible to use multiple
  times. An term which is central, relevant, and affine is called
  \textit{pure}. The rules for this judgement are given in Figure
  \ref{fig:ssa-term-typing}.

@@

We can now get to the typing judgement for expressions in Figure
\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
for variables, with the requirement that, other than \(x\), all variables in
\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{term}
\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
reflects that we're allowed to \textit{forget} capabilities, but not
\textit{add} them. The \rle{app} rule is also mostly standard: given an
instruction \(f\) which takes in an argument \(A\) and has a return type of
\(B\) and a \textit{central} term \(\hasty{\Gamma}{1}{a}{A}{q}\), we
conclude that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that
\(f\) itself have centrality \(p\) and quantity \(q\), i.e. that \(f \in
\mc{I}_p^q(A, B)\). The \rle{pair} rule says that we can type a pair \((a, b)\)
in \(\Gamma\) if \(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\)
and \(\Xi\) types \(b\); the quantity of a pair is simply the intersection of
the quantity of its components (this is expressed in the rule by allowing any
quantity which types \textit{both} components), and both components must be
central. Finally, we have the rules \rle{let} and \rle{let2}, corresponding to
let-bindings and destructuring let-bindings respectively. In both cases, we
split the context \(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is
used to type a central subterm \(a\), which is bound to variables \(x\) or
destructured to variables \(x, y\) respectively, in a potentially non-central
term \(e\) which is otherwise typed in \(\Delta\).

@@

The rules for block typing are given in Figure \ref{fig:ssa-block-typing}.
\rle{br} says that, given a pure term \(a: A\) (that is, central, relevant, and
affine), we can type a block \(\lbrb{\ell}{a}\) which branches to \(\ell\) with
argument \(a\), where \(\ms{L}\) contains the context \(\ell\) with an argument
type of \(A\) and an appropriate set of unused variables. The rule \rle{ite}
typechecks if-then-else statements by stating that, if \(\Gamma\) splits into
\(\Delta\) and \(\Xi\), where \(\Xi\) typechecks a pure, intuitionistic
discriminant \(e: \mb{2}\) and \(\Delta\) typechecks both \(s, t\) to branch
into \(\ms{L}\), then the whole if-statement \(\ite{e}{s}{t}\) also typechecks
as branching into \(\ms{L}\) in constext \(\Gamma\). \rle{let-blk} and
\rle{let2-blk} are quite similar to their term analogues, except that the
bound term \(e\) is allowed to be impure (otherwise, programs could only
make use of pure functions!), and the remainder of the context plus the bound
variable is used to type a block rather than a term. Finally, probably the most
interesting rule is \rle{where}, which says that, given a block \(t\) which
branches to \(\ms{L}\) in context \(\Gamma\), and labels \(L\) mapping
\(\ms{L}\) to \(\ms{K}\), \(\ewhere{t}{L}\) branches to \(\ms{K}\). Here, \(L\)
mapping \(\ms{L}\) to \(\ms{K}\) is defined using the typing rules \rle{nil-br},
which says that an empty control-flow graph simply forwards the input labels
directly to the output labels, and \rle{cons-br}, which says that if \(L\) maps
\(\ms{L}\) to either \(\ms{K}\) or the label \(\lbl{\ell}[\Gamma](A)\), and
\(t\) branches to \(\ms{L}\) in the context \(\Gamma, \thyp{x}{A}{}\), then
\(L\) maps \(\ms{L}\) to \(\ms{K}\), having (potentially recursively) defined
\(\lbl{\ell}\).

@@

Given such a substitution \(\gamma\), there are two ways to use it. The first is
standard capture-avoiding substitution, which we denote \([\gamma]e\) for
expressions, \([\gamma]t\) for blocks, and \([\gamma]L\) for CFGs, as is
standard. The second is to use the substitution to generate a set of
let-bindings for a context \(\Gamma\), written
\(\exprletsubst{\substctx{\gamma}{\Gamma}}{e}\) for expressions and
\(\stmtletsubst{\substctx{\gamma}{\Gamma}}{t}\) for blocks, as defined in Figure
\ref{fig:ssa-subst-ops}. Note that this operation cannot be easily defined for
CFGs. In Section \ref{sec:semantics}, we will show that this is semantically
equivalent to regular capture-avoiding substitution, due to the fact that all
terms in a substitution are central \textit{and} that only affine computations
can be used in a term substituting for an affine variable, and only relevant
computationbs can be used in a term substituting for a relevant variable.

@@

We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
relevant term becomes a relevant morphism, and so, in particular, a pure term
becomes a pure morphsim. Our rules for doing so are given in Figure
\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
\(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
"judgement"; this is justified by the coherence lemma:
\begin{lemma}[Coherence] \
  \begin{itemize}
    \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
    \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
  \end{itemize}
\end{lemma}
Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
if all other variables in \(\Gamma\) and can be discarded, in which case the
semantics is simply doing so, i.e. \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\),
lifting to \(\mc{C}_0\) as necessary. Similarly, to type a function application
\rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is
guarnateed to be central, lift it to the appropriate category \(\mc{C}_p\), and
then simply postcompose it with the denotation of \(f\). \rle{pair} is handled
by composing the context split \(\csplits{\Gamma}{\Delta}{\Xi}\) taking the
tensor product of the components \(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\),
\(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is well-defined since both components
are central and \(\mc{C}_1\) is monoidal. We then lift the result to the
appropriate \(\mc{C}_p\). The semantics of \rle{unit} is simply dropping the
entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\), while \rle{true}, and \rle{false}
are each handled by first dropping the context and then postcomposing with the
appropriate constant; in all cases, this is followed by lifting to \(\mc{C}_p\).
Finally, \rle{let} and \rle{let2} are handled by first splitting the context,
then passing the right hand component to the semantics of the bound term
\(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or \(\dnt{\hasty{\Xi}{1}{a}{A \otimes
B}{q}}\), and then, lifting to \(\mc{C}_p\) as necessary, after reassociating
with \(\alpha\), passing everything to the inner term \(\dnt{\hasty{\Delta,
\thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
\thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that this inner term is not necessarily
central, unlike the bound term!

@@

Similarly to terms, \textit{blocks} \(\haslb{\Gamma}{t}{\ms{L}}\) are
interpreted as \textit{impure} morphisms from \(\dnt{\Gamma}\) to \(\ms{L}\):
here \(\Gamma\) represents all the variables live on entry to the block, while
\(\ms{L}\) represents all the external blocks which \(t\) might branch to, each
annotated with their live variables and their arguments; our rules are given in
Figure \ref{fig:ssa-block-semantics}. Our semantics for break statements
\(\lbrb{\ell}{a}\) is as follows: we begin by splitting the context into live
variables which pass into dominated blocks and variables used in the argument
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\). We then pass the latter into the
semantics of the argument (which is pure, and therefore must be lifted to
\(\mc{C}_0\)) \(\dnt{\hasty{\Gamma}{1}{a}{A}{\tint}}\), with the result and live
variables re-associated and passed into
\(\lwk{\llhyp{\ell}{\Delta}{A}}{\ms{L}}\) to obtain the desired target. The
semantics of if-statements are similarly given by first splitting the context
via \(\dnt{\csplits{\Gamma}{\Delta}{\Xi}}\) into variables live in the branches,
\(\Delta\), and variables appearing in the discriminant, \(\Xi\); the latter of
which are passed into the lifted pure morphism
\(\dnt{\hasty{\Delta}{1}{b}{\mb{2}}{\tint}}\). The "if" statement itself is then
simply an application of the distributor \(\delta: A \otimes (\mb{1} + \mb{1})
\to A + A\), with each branch passed to the semantics of its corresponding block
\(\dnt{\haslb{\Xi}{s}{\ms{L}}}\), \(\dnt{\haslb{\Xi}{t}{\ms{L}}}\). The
semantics of \rle{let-blk} and \rle{let2-blk} are analogous to those of
\rle{let} and \rle{let2} in the term case, except that now the bound term \(a\)
is allowed to be impure (and therefore in particular does not need to be
upgraded).

@@

Because we have a denotational semantics, it is easy to validate equations
between terms, blocks, and CFGs.

@@

To properly be able to reason about congruence and congruence-based
optimization, we need to introduce the concept of \textit{terms with holes},
which are defined using the grammar in Figure
\ref{fig:blocks-with-holes-grammar}. This simply extends the three main
syntactic categories with \textit{holes} of the form \(\lhole{X}\),
\(\lhole{T}\), \(\lhole{W}\). To \textit{type} terms with holes, we need to
introduce the concept of a \textit{hole-context} \(H\), which types a hole as being
one of:
\begin{itemize}
  \item \todo{term}
  \item \todo{block}
  \item \todo{control flow graph}
\end{itemize}
\TODO{hole typing rules for expressions, blocks, control-flow graphs; everything
else is the same}.
